Quantifying Directionality and Innovation Output in Sweden's Transition to a Forest-Based Bioeconomy

### 2024-05-06 paper.qmd
update paper for midway draft

csl: 'https://zotero.org/styles/apa?source=1'
bibliography:
  - swinno-shades-green.bib
link-citations: true
link-bibliography: true
colorlinks: true
number-sections: true
format:
  pdf:
    documentclass: scrartcl
    papersize: a4
    linestretch: 1.25
    mainfont: "Adobe Garamond Pro"
    sansfont: "Adobe Garamond Pro Semibold"
execute:
  warning: false
  error: false
  echo: false
jupyter: python3
---

@@

<!-- Paragraphs 1-2: Motivation. After reading these paragraphs a reader in any field of economics should believe that if you answer your research question your paper will make an important contribution. -->
At the heart of sustainability transitions lies the question how societal functions can be fulfilled within ecological boundaries [@kohler2019AgendaSustainabilityTransitions].
Societal functions are provided through socio-technical systems, which consist of the interplay between the production, distribution, and use of technology [@geels2004SectoralSystemsInnovation].
Transitions are complex, long-term processes of shifts between different systems, involving diverse actors who have to chose between multiple, open-ended pathways, uncertain of their technological, social and political success [@kohler2019AgendaSustainabilityTransitions].
Further, success in the outcome and chosen pathway of sustainability transitions is contested based on the values, norms, and goals of  different actors and social groups.
Hence, innovation cannot be seen as a scalar, but must rather be understood as a vector, as possessing and providing direction to change rather than being simply progress forward [@stirling2011PluralisingProgressIntegrative].
@andersson2021OutcomesDirectionalityMorphology term these two properties _positive_ and _normative_ directionality, respectively.
Positive directionality of innovation arises from the fact that technological change is path-dependent [@arthur1989CompetingTechnologiesIncreasing, @david1985ClioEconomicsQWERTY], while normative directionality is rooted in the assumption that without public policy, private actors lack intrinsic incentives to pursue sustainability [@kohler2019AgendaSustainabilityTransitions].

A sustainability transition which has been gaining relevance for researchers, policy makers, and producers alike is the bioeconomy.
Broadly speaking, it aims to radically shift economic systems from relying on fossil fuels for energy and material resource, to an economic system built on biobased materials [@bennich2018BioBasedEconomyDynamics], circular and renewable value chains [@damato2017GreenCircularBio] and higher environmental integrity through innovation [@bugge2016WhatBioeconomyReview].
As one of the main sources for biomass in a bioeconomy [@priefer2017PathwaysShapeBioeconomy], forests have become a key arena for the bioeconomy transition, highlighting tradeoffs between material provision, carbon storage and biodiversity [@hogbom2021TrilemmaNordicBaltic].
Hence, underneath the bioeconomy umbrella, the concept and viable transition pathways are heavily contested [@allain2022BioeconomicsVsBioeconomy].
Three different visions of a bioeconomy compete in discourse to provide normative directionality of the transition; the first vision centers on the provision of biomass and renewable resources, the second on the scientific development and application of biotechnology, while the third emphasizes ecological considerations[@bugge2016WhatBioeconomyReview; @vivien2019HijackingBioeconomy].
While the discourse is well understood in terms of content [e.g., @bugge2016WhatBioeconomyReview, @vivien2019HijackingBioeconomy] its narratives [@bauer2018NarrativesBiorefineryInnovation], and actors [@holmgren2022WhoseTransformationThis], the material effects of this transition, however, remain important research topics [@fischer2020SwedishForestsBioeconomy].
How many innovations are commercialized in this new economy, especially considering the central role of innovation in transition discourse? And what does commercialized innovation suggest for the positive directionality of the bioeconomy transition?

@@

To unpack the plurality of possible socio-technical development paths, places high demands on the data quality of innovation measurement.
Although progress towards monitoring the bioeconomy transition has been made [@jander2019MonitoringTransitionBioeconomy; @jander2020MonitoringBioeconomyTransitions], assessing directionality is limited by a lack of innovation output measures [@wydra2020MeasuringInnovationBioeconomy].
In this paper, I use SWINNO, a literature-based innovation output database capturing all significant commercialized innovation in Sweden [@sjoo2014DatabaseSwedishInnovations], to analyze the innovation output of the Swedish forest-based bioeconomy.
I develop and apply a classification scheme to identify which bioeconomy vision an innovation produced or used in forestry or the downstream value chain aligns with.

@@

Path-dependency arises from events at one point in time affect later events, and may even lead to suboptimal outcomes [@david1985ClioEconomicsQWERTY].
It arises due to positive feedback effects, such as increasing returns [@arthur1989CompetingTechnologiesIncreasing], or through extension of possibilities adjecent to exiting technology [@kauffman2000Investigations].
Hence, technological change is contingent and influences technological change at previous or later points in time.
Instead of more or less progress, each innovation implies a new direction for socio-technical systems [@stirling2011PluralisingProgressIntegrative].
Andersson et al., [-@andersson2021OutcomesDirectionalityMorphology] term this quality of innovation possessing direction _positive_ directionality.

@@

Dealing with the question of how to provide and exercise normative transition directionality @weber2012LegitimizingResearchTechnology synthesis policy insights from innovation systems and multilevel perspective research into a failures framework.
Therein, directionality failure is understood as the failure to provide direction to the process of search and unclear or missing setting of collective priorities.
They caution that it should not be confused with shortsighted or flawed anticipation of market developments, as this myopia lacks a view on the direction of change.
To overcome this failure, that is, to provide directionality, they argue, guiding orientation needs to be translated and intermediated on two levels.
External requirements placed on the innovation system need to be absorbed after which they need to be interpreted in a process of negotiation between the different actors.
Reaching this collective understanding of direction, @weber2012LegitimizingResearchTechnology [p. 1042] write, is in practice, "achieved by establishing shared future visions".

In the two central research streams of sustainability transitions, the innovation systems and the multi-level perspective literature, the importance of visions is highlighted.
@hekkert2007FunctionsInnovationSystems provide a framework of intervention in innovation on a systemic level, which highlights the importance of guidance of search as one of seven functions innovation system must provide.
They explain guidance of search as, "those activities within the innovation system that can positively affect the visibility and clarity of specific wants among technology users" [@hekkert2007FunctionsInnovationSystems, 423].
It is changing preferences in society, which through market and government influence and, ".. an interactive and cumulative process of exchanging ideas between technology producers, technology users, and many other actors" [@hekkert2007FunctionsInnovationSystems, 423] decides which direction of technological development is most promising. 
@bauer2017TechnologicalInnovationSystems adapts the innovation system functions to the specific case of biorefineries in Sweden and define the influence on the direction of search as "the incentives and/or pressures for organizations to enter the technological field. These may come in the form of visions, expectations of growth potential, regulation, policy targets, standards, articulation of demand from leading customers, crises in current business, etc." (p. 4).
Using the multi-level perspective, @yang2022ShapingDirectionalitySustainability too highlight the importance of visions in directionality over transformations, stressing the importance of shared visions between niche and regime actors.
Similarly, @bolton2019GrandVisionsPragmatic illustrate how visions and pragmatic change combined in the development and evolution of a highly integrated European electricity system.

@rotmans2001MoreEvolutionRevolution even go so far to argue that transition management is based on appealing and imaginative long-term visions and @loorbach2010TransitionManagementSustainable argues that shared visions need to be developed and disseminated between transition stakeholders.
But, as @bolton2019GrandVisionsPragmatic show, visions alone are insufficient to provide directionality.
In addition to being appealing, visions must also be specific and actors need to buy into them [@grillitsch2019InnovationPolicySystemwide].
In summary, visions can be seen as shared imaginaries, mental models and aspirations of desired transitions which are expressed through narratives in discourse between actors [@bauer2018NarrativesBiorefineryInnovation; @holmgren2022WhoseTransformationThis].

@@

Although the bioecononmy has gained importance as a research, policy and industry focus, it remains unclear what exactly it entails.
At a high level, the bioeconomy refers to an economy in which raw materials for energy and production processes are provided from biological instead of fossil sources [@bugge2016WhatBioeconomyReview].
Hopes of increased resource efficiency and renewability as well as promised economic growth especially in regions rich in biological resources abound [@el-chichakli2016PolicyFiveCornerstones].
However, a such a profound change to economic systems will not be possible without balancing significant tradeoffs between the needs and desires of different human and non-human stakeholders [@dieken2021MultitudesBioeconomiesSystematic]. @hogbom2021TrilemmaNordicBaltic summarize these tradeoffs as a trilemma between forests ability to provide sufficent wood to replace fossil based materials, carbon sequestration and storage, and the delivery of other eco-system services such as forest biodiversity.
Resolving these tradeoffs is challenging, as goal conflicts between different development pathways make simulatneous progress on social, ecological and economic goals difficult [@bennich2021BiobasedEconomy2030].

As one of the leading forestry nations in the world, Sweden sees substantial economic and political potential in advancing the bioeconomy .
In a report commissioned by the Swedish Government, the @2012StrategyBiobasedEconomy lays out the plan for a Swedish bioeconomy based on two central pillars.
It is to be an economy based on,

@@

These pillars reflect the two most influential visions of the bioeconomy concept. Notions of a biotechnology and bio-resource oriented bioeconomy are widespread, dominate discourse at various levels [@dieken2021MultitudesBioeconomiesSystematic] and form the main industrial visions [@befort2020GoingDefinitionsUnderstand]. 
A third vision, centered around ecology and an economy compatible with the biosphere acts as a counter point.

@@

The application of biotechnology to promote economic growth and provide jobs lends the biotechnology vision its name [@bugge2016WhatBioeconomyReview]. Science driven innovation features prominently in this vision, with collaborations between research institutions in academia and industry taking on an important role [@bugge2016WhatBioeconomyReview]. @vivien2019HijackingBioeconomy present a review of the concept showing that the focus on science as an absolute motor of change also shaped policy recommendations, driven especially by the OECD, towards increasing the speed and diffusion of innovation through partnerships between new start-ups and establish pharmaceutical companies. As a policy narrative this vision was particularly potent and widespread at the turn of the 21st century. According to @vivien2019HijackingBioeconomy, the intellectual heritage of Schumpeter and Kontradieff play an important role in this vision. Technological breakthroughs are poised to solve economic and ecological challenges, for example through genetic engineering. Consequently, this vision has a weak sustainability notion, viewing sustainability challenges as temporary, and nature as substitutable given sufficient research and knowledge creation [@vivien2019HijackingBioeconomy].

In more recent years, visions of a bio-resource bioeconomy have taken over as the leading narrative [@allain2022BioeconomicsVsBioeconomy]. Unlike the biotechnology vision, this conceptualization does not only focus on economic growth and job creation, but attempts to include ecological sustainability challenges [@bugge2016WhatBioeconomyReview]. Biological resources, not technologies, form the basis of a new economy. In essence, there are two ways in which bio-resources are poised to replace fossil resources: as substitution, e.g., biofuels replacing fossil fuels, and as high value added products, e.g., cross-laminated timber used as an alternative to concrete in construction [@vivien2019HijackingBioeconomy]. To substitute fossil resources requires significant efforts in innovation. In contrast to the biotechnology vision, however, innovation is not confined to scientific advances from a narrow biochemical and pharmaceutical field. Instead, new and old actors are required to collaborate between and beyond established sectors [@bugge2016WhatBioeconomyReview]. Although sustainability plays a more important role in this vision of the bioeconomy, this vision is considered an example of weak sustainability [@damato2017GreenCircularBio; @vivien2019HijackingBioeconomy]. The reason is the focus of substitution and high biomass demands. The consumption of raw materials may even increase, not only for bio-based, but also for fossil-based resources [@asada2020BioeconomicTransitionProjecting].

The exact role played by forests in a bioeconomy transition depends on the desired vision [@kleinschmit2014ShadesGreenSocial]. There is no doubt that the Swedish vision places high demands on forests as primary providers of the required biomass. Or, as @belandlindahl2015SwedishForestryModel call it a forest sector that operates under a "more of everything" paradigm. Both in the strategy formulated in @research2012StrategyBiobasedEconomy and in public discourse, a small, well-connected group of actors center the bio-resource vision as legitimate and desirable [@holmgren2022WhoseTransformationThis].

A paragraph here explaining the detailed findings of @holmgren2022WhoseTransformationThis discourse network here.

Despite -- or perhaps because of -- bioeconomy lacking a strong definition, the term is used to refer to different concepts, with different aims and objectives, foci, normative assumptions and hence implications for society and environment [@bugge2016WhatBioeconomyReview].
Common ground in ambitions to realize a bioeconomy is the need for diverse types of social, technological and organizational innovation [@el-chichakli2016PolicyFiveCornerstones].  According to @jankovsky2021InnovationsForestBioeconomy there are four types of innovation crucial for a forest-based bioeconomy: (i) substitute products, (ii) new (bio-based) processes, (iii) new (bio-based) products, (iv) new behavior.
The need for innovation also features large in Swedish bioeconomy discourse and narratives therein [@bauer2017TechnologicalInnovationSystems, @bennich2021BiobasedEconomy2030], as, "\[t\]echnological innovation is key to a greener future" [@holmgren2022WhoseTransformationThis 42].

For the remainder of the paper the nomenclature of @bugge2016WhatBioeconomyReview is adopted.
Vivien et al.'s [-@vivien2019HijackingBioeconomy] Type I bioeconomy is taken to correspond with the bioecology vision, Type II and Type III to correspond with the bioresource and biotechnology vision, respectively.
The analysis of bioeconomy discourse which revealed different visions, highlights that the visions have emerged at different times and with different actors at their centers [@vivien2019HijackingBioeconomy, @bugge2016WhatBioeconomyReview].
@vivien2019HijackingBioeconomy trace the introduction of the term bioeconomics to the 1920s and a work on fishery by the Russian biologist Baranoff.
The term bioeoconomics which they label as Type I bioeconomy, was introduced by Georgescu-Roegen, in -@georgescu-roegen1975BioeconomicAspectsEntropy.
@vivien2019HijackingBioeconomy argue that the term biotechnology gradually emerged between 1990 and 2000, in the wake of enthusiasm about biotechnological advances, such as the discovery of DNA.
It gained popularity and international importance when the OECD began increasingly supporting and reshaping its meaning at the turn of the millennium [@vivien2019HijackingBioeconomy].
According to them, the last concept to enter the discourse was the bioresource oriented bioeconomy.
While biomass has been used throughout history in various ways, the renewed interest with the intention to replace fossil inputs, for example in European Commission communication since 2010.
However, @vivien2019HijackingBioeconomy point to the long history of biorefinery as a core aspect of this vision, which was discussed as early as the 1930s.
Hence, while the organization of the discourse into the labels biotechnology, bioecology and bioresource vision is a very recent research outcome, the underlying concepts have a richer history.

@@

To quantify transition directionality, data about innovation activity is needed.
Operationalizing directionality hence difficult.
Few notable attempts such as
@cappellano2022WhereInnovationMeets operationalize directionality as a measure of R&I expenditure earmarked for specific transition objectives.
Other attempts to measure the directionality and progress towards a bioeconomy @jander2019MonitoringTransitionBioeconomy and @jander2020MonitoringBioeconomyTransitions, but lack the plurality of innovation.
Previous research has frequently relied on input and throughput indicators of innovation activities, such as R&D expenditure, or  publications and patents [@wydra2020MeasuringInnovationBioeconomy].
Such data, however, suffers from limitations when attempting to assess the directionality of socio-technical systems.
While R&D expenditure is a key lever to exercise normative directionality for policy makers, the uncertainty of innovation requires portfolio diversification to prevent misallocation of public and private resources.
Patents on the other hand, suffer from well documented measurement errors; not all innovations are patented, and not all patents become innovations [@arundel1998WhatPercentageInnovations].

Literature-based innovation output measures remedy these issues, by capturing commercialized innovations [@vanderpanne2007IssuesMeasuringInnovation].
Constructed with this method in Sweden, the SWINNO database contains `{python} n_swinno` innovations commercialized by Swedish companies between `{python} year_min` and `{python} year_max` [@sjoo2014DatabaseSwedishInnovations].
SWINNO is constructed from editor-reviewed articles published in 15 independent trade journals covering technological developments for general and highly specialized audiences [@sjoo2014DatabaseSwedishInnovations].
Since this method relies on the judgment of editiors and journalists, it can be described as significant innovations, omitting small reconfigurations of exisiting products and processes.
The SWINNO database has significant benefits compared to other innovation measures.
For one, it ensures that actual innovation, meaning commercialized products and processes, and thus that its positive directionality is captured.
Second, because the source articles are preserved, it provides a rich source of data, giving frequently providing insights into the aims, motivations, advantages and challenges of the covered innovation, allowing to assess and quantify the vector of technological change.

To capture the forest-based bioeconomy the database was searched for innovation associated with core forest sectors and keywords describing forest products which may be used in sectors along the bioeconomy value chain based on @wolfslehner2016ForestBioeconomyNew.
The list of keywords and sectors is presented in the appendix.

@@

Querying the SWINNO database for forest based bioeconomy innovations initially yields `{python} n_uncleaned_bioeconomy_innovations` innovations. This number includes false positives from substrings matching keywords included in the search. During classification, these false positives were removed from the dataset, yielding `{python} sum_bioeconomy_innovations` innovations which originate from, are used in or make use of products derived from the forest.

@@

For the following part, innovations which do not relate to the forest are excluded. Excluded are also the `{python} n_uncertain_bioeconomy_visions` innovations whose source text did not allow labeling without significant uncertaintiy. These innovations are planned to undergo a revision by subject matter experts at a later point.

>>>>

csl: 'https://zotero.org/styles/apa?source=1'
bibliography:
  - swinno-shades-green.bib
link-citations: true
link-bibliography: true
colorlinks: true
number-sections: true
number-depth: 3
format:
  pdf:
    documentclass: scrartcl
    papersize: a4
    linestretch: 1.25
    mainfont: "Adobe Garamond Pro"
    sansfont: "Adobe Garamond Pro Semibold"
    indent: true
execute:
  warning: false
  error: false
  echo: false
jupyter: python3
---

@@

<!-- Paragraphs 1-2: Motivation. After reading these paragraphs a reader in any field of economics should believe that if you answer your research question your paper will make an important contribution. -->
At the heart of sustainability transitions lies the question how societal functions can be fulfilled within ecological boundaries [@kohler2019AgendaSustainabilityTransitions].
Societal functions are provided through socio-technical systems, which consist of the interplay between the production, distribution, and use of technology [@geels2004SectoralSystemsInnovation].
Transitions are complex, long-term processes of shifts between different systems, involving diverse actors who have to chose between multiple, open-ended pathways, uncertain of their technological, social and political success [@kohler2019AgendaSustainabilityTransitions].
Further, success in the outcome and chosen pathway of sustainability transitions is contested based on the values, norms, and goals of  different actors and social groups.
Hence, innovation cannot be seen as a scalar, but must rather be understood as a vector, as possessing and providing direction to change rather than being simply progress forward [@stirling2011PluralisingProgressIntegrative].
@andersson2021OutcomesDirectionalityMorphology term these two properties _positive_ and _normative_ directionality, respectively.
Positive directionality of innovation arises from the fact that technological change is path-dependent [@arthur1989CompetingTechnologiesIncreasing;@david1985ClioEconomicsQWERTY], while normative directionality is rooted in the assumption that without public policy, private actors lack intrinsic incentives to pursue sustainability [@kohler2019AgendaSustainabilityTransitions].

A sustainability transition which has been gaining relevance for researchers, policy makers, and producers alike is the bioeconomy.
Broadly speaking, it aims to radically shift economic systems from relying on fossil fuels for energy and material resource, to an economic system built on biobased materials [@bennich2018BioBasedEconomyDynamics], circular and renewable value chains [@damato2017GreenCircularBio] and higher environmental integrity through innovation [@bugge2016WhatBioeconomyReview].
As one of the main sources for biomass in a bioeconomy [@priefer2017PathwaysShapeBioeconomy], forests have become a key arena for the bioeconomy transition, highlighting tradeoffs between material provision, carbon storage and biodiversity [@hogbom2021TrilemmaNordicBaltic].
Underneath the umbrella term "bioeconomy", the concept and viable transition pathways are heavily contested [@allain2022BioeconomicsVsBioeconomy].
Three different visions of a bioeconomy compete in discourse to provide normative directionality of the transition: the first vision centers on the provision of biomass and renewable resources, the second on the scientific development and application of biotechnology, while the third emphasizes ecological considerations[@bugge2016WhatBioeconomyReview; @vivien2019HijackingBioeconomy].
Although the discourse is well understood in terms of content [e.g., @bugge2016WhatBioeconomyReview;@vivien2019HijackingBioeconomy] its narratives [@bauer2018NarrativesBiorefineryInnovation], and actors [@holmgren2022WhoseTransformationThis], the material effects of this transition, however, remain important research topics [@fischer2020SwedishForestsBioeconomy].
How many innovations are commercialized in this new economy, especially considering the central role of innovation in bioeconomy discourse? And what does commercialized innovation suggest for the positive directionality of the bioeconomy transition?

@@

To unpack the plurality of possible socio-technical development paths, requires high data quality when measuring innovation.
Although progress towards monitoring the bioeconomy transition has been made [@jander2019MonitoringTransitionBioeconomy; @jander2020MonitoringBioeconomyTransitions], assessing directionality is limited by a lack of innovation output measures [@wydra2020MeasuringInnovationBioeconomy].
In this paper, I use SWINNO, a literature-based innovation output database capturing all significant commercialized innovation in Sweden [@sjoo2014DatabaseSwedishInnovations], to analyze the innovation output of the Swedish forest-based bioeconomy.
I develop and apply a classification scheme to identify which bioeconomy vision an innovation produced or used in forestry or the downstream value chain aligns with.

@@

Technology does not change randomly.
Instead, technological change is contingent and influences technological change at previous or later points in time.
It is path-dependent, meaning events at one point in time affect later events, which may even lead to suboptimal outcomes [@david1985ClioEconomicsQWERTY].
Path-dependency arises due to positive feedback effects, such as increasing returns [@arthur1989CompetingTechnologiesIncreasing], or through extension of possibilities adjacent to exiting technology [@kauffman2000Investigations].
Instead of more or less progress, each innovation implies a new direction for socio-technical systems [@stirling2011PluralisingProgressIntegrative].
@andersson2021OutcomesDirectionalityMorphology term this quality of innovation possessing direction _positive_ directionality.

@@

Dealing with the question of how to provide and exercise normative transition directionality  Weber and Rohracher [-@weber2012LegitimizingResearchTechnology] synthesize policy insights from innovation systems and multilevel perspective research into a failures framework.
Therein, directionality failure is understood as the failure to provide direction to the process of search and unclear or missing setting of collective priorities.
They caution that it should not be confused with shortsighted or flawed anticipation of market developments, as this myopia lacks a view on the direction of change.
To overcome this failure, that is, to provide directionality, they argue, external requirements need to be internalized by the innovation system in a interpretive process of negotiation between different system actors.
Reaching this collective understanding of direction, Weber and Rohracher [-@weber2012LegitimizingResearchTechnology p. 1042] write, is in practice, "... achieved by establishing shared future visions".

In the two central research streams of sustainability transitions, the innovation systems and the multi-level perspective literature, the importance of visions is highlighted.
@hekkert2007FunctionsInnovationSystems provide a framework of intervention in innovation on a systemic level, which highlights the importance of guidance of search as one of seven functions innovation system must provide.
They explain guidance of search as, "those activities within the innovation system that can positively affect the visibility and clarity of specific wants among technology users" [@hekkert2007FunctionsInnovationSystems, 423].
Changing preferences in society, decides which direction of technological development is most promising through market and government influence and, ".. an interactive and cumulative process of exchanging ideas between technology producers, technology users, and many other actors" [@hekkert2007FunctionsInnovationSystems, 423]. 
@bauer2017TechnologicalInnovationSystems adapts the innovation system functions to the specific case of biorefineries in Sweden and define the influence on the direction of search as "the incentives and/or pressures for organizations to enter the technological field. These may come in the form of visions, expectations of growth potential, regulation, policy targets, standards, articulation of demand from leading customers, crises in current business, etc." (p. 4).
Using the multi-level perspective, @yang2022ShapingDirectionalitySustainability too highlight the importance of visions in directionality over transformations, stressing the importance of shared visions between niche and regime actors.
Similarly, @bolton2019GrandVisionsPragmatic illustrate how visions and pragmatic change combined in the development and evolution of a highly integrated European electricity system.

@rotmans2001MoreEvolutionRevolution even go so far to argue that transition management is based on appealing and imaginative long-term visions and @loorbach2010TransitionManagementSustainable argues that shared visions need to be developed and disseminated between transition stakeholders.
But, as @bolton2019GrandVisionsPragmatic show, visions alone are insufficient to provide directionality.
In addition to being appealing, visions must also be specific and actors need to buy into them [@grillitsch2019InnovationPolicySystemwidea].
In summary, visions can be seen as shared imaginaries, mental models and aspirations of desired transitions which are expressed through narratives in discourse between actors [@bauer2018NarrativesBiorefineryInnovation; @holmgren2022WhoseTransformationThis].

@@

Although the bioeconomy has gained importance as a research, policy and industry focus, it remains unclear what exactly it entails.
At a high level, the bioeconomy refers to an economy in which raw materials for energy and production processes are provided from biological instead of fossil sources [@bugge2016WhatBioeconomyReview].
Hopes of increased resource efficiency and renewability as well as promised economic growth especially in regions rich in biological resources abound [@el-chichakli2016PolicyFiveCornerstones].
However, a such a profound change to economic systems will not be possible without balancing significant tradeoffs between the needs and desires of different human and non-human stakeholders [@dieken2021MultitudesBioeconomiesSystematic]. @hogbom2021TrilemmaNordicBaltic summarize these tradeoffs as a trilemma between forests ability to provide sufficent wood to replace fossil based materials, carbon sequestration and storage, and the delivery of other eco-system services such as forest biodiversity.
Resolving these tradeoffs is challenging, as goal conflicts between different development pathways make simulatneous progress on social, ecological and economic goals difficult [@bennich2021BiobasedEconomy2030].

As one of the leading forestry nations in the world, Sweden sees substantial economic and political potential in advancing the bioeconomy .
In a report commissioned by the Swedish Government, @formas2012SwedishResearchInnovation lays out the plan for a Swedish bioeconomy based on two central pillars.
It is to be an economy based on,

@@

While the official Swedish bioeconomy strategy is still under development, the content of the quotes above echos across policy documents and the national forest program [@fischer2020SwedishForestsBioeconomy].
These pillars reflect the two most influential visions of the bioeconomy concept. Notions of a biotechnology and bio-resource oriented bioeconomy are widespread, dominate discourse at various levels [@dieken2021MultitudesBioeconomiesSystematic] and form the main industrial visions [@befort2020GoingDefinitionsUnderstand]. 
A third vision, centered around ecology and an economy compatible with the biosphere acts as a counter point.

@@

The application of biotechnology to promote economic growth and provide jobs lends the biotechnology vision its name [@bugge2016WhatBioeconomyReview]. 
Biotechnology is the application of cellular and / or molecular biological agents in the production or processing of materials in a scientific and engineering informed manner [@yeung2019CurrentResearchBiotechnology].
Science driven innovation features prominently in this vision, with collaborations between research institutions in academia and industry taking on an important role [@bugge2016WhatBioeconomyReview]. @vivien2019HijackingBioeconomy present a review of the concept showing that the focus on science as an absolute motor of change also shaped policy recommendations, driven especially by the OECD, towards increasing the speed and diffusion of innovation through partnerships between new start-ups and establish pharmaceutical companies. As a policy narrative this vision was particularly potent and widespread at the turn of the 21st century. According to @vivien2019HijackingBioeconomy, the intellectual heritage of Schumpeter and Kontradieff play an important role in this vision. Technological breakthroughs are poised to solve economic and ecological challenges, for example through genetic engineering. Consequently, this vision has a weak sustainability notion, viewing sustainability challenges as temporary, and nature as substitutable given sufficient research and knowledge creation [@vivien2019HijackingBioeconomy].

In more recent years, visions of a bio-resource bioeconomy have taken over as the leading narrative [@allain2022BioeconomicsVsBioeconomy]. Unlike the biotechnology vision, this conceptualization does not only focus on economic growth and job creation, but attempts to include ecological sustainability challenges [@bugge2016WhatBioeconomyReview]. Biological resources, not technologies, form the basis of a new economy. In essence, there are two ways in which bio-resources are poised to replace fossil resources: as substitution, e.g., biofuels replacing fossil fuels, and as high value added products, e.g., cross-laminated timber used as an alternative to concrete in construction [@vivien2019HijackingBioeconomy]. To substitute fossil resources requires significant efforts in innovation. In contrast to the biotechnology vision, however, innovation is not confined to scientific advances from a narrow biochemical and pharmaceutical field. Instead, new and old actors are required to collaborate between and beyond established sectors [@bugge2016WhatBioeconomyReview]. Although sustainability plays a more important role in this vision of the bioeconomy, this vision is also considered an example of weak sustainability [@damato2017GreenCircularBio; @vivien2019HijackingBioeconomy]. The reason is the focus of substitution and high biomass demands. The consumption of raw materials may even increase, not only for bio-based, but also for fossil-based resources [@asada2020BioeconomicTransitionProjecting].

An ecology focused counter-vision exists, which @bugge2016WhatBioeconomyReview term the bioecology vision, while ecological economists preferentially use the term bioeconomics [@vivien2019HijackingBioeconomy; @allain2022BioeconomicsVsBioeconomy].
Unlike the previous two visions, the bioecology vision builds on strong sustainability assumptions and emphasizes the need to develop an alternative economic system which respects the limits of the biosphere.
Normative change towards material sobriety and degrowth to protect and restore ecosystems and biodiversity take precedent over technological and economic ambitions [@vivien2019HijackingBioeconomy;@bugge2016WhatBioeconomyReview].

#### Development of the Term Bioeconomy {#sec-bioeconomy-term-development}

Analysis of the bioeconomy discourse, highlights that the visions have emerged at different times and with different actors at their centers [@vivien2019HijackingBioeconomy; @bugge2016WhatBioeconomyReview].
And while the two hitherto mentioned visions dominate contemporary discourse towards a forest based bioeconomy [@holmgren2022WhoseTransformationThis; @bauer2018NarrativesBiorefineryInnovation], the bioeconomics concept itself was introduced in the 1920s in a work on fishery by the Russian biologist Baranoff [@vivien2019HijackingBioeconomy].
Later, in 1975 ecological economist Georgescu-Roegen [-@georgescu-roegen1975BioeconomicAspectsEntropy] used the term to argue for an economy which minimizes its material and energy throughput to respect biological, physical and geochemical limits.
@vivien2019HijackingBioeconomy argue that the term biotechnology gradually emerged between 1990 and 2000, in the wake of enthusiasm about biotechnological advances, such as the discovery of DNA.
It gained popularity and international importance when the OECD began increasingly supporting and reshaping its meaning at the turn of the millennium [@vivien2019HijackingBioeconomy].
According to them, the last concept to enter the discourse was the bioresource oriented bioeconomy.
While biomass has been used throughout history in various ways, the renewed interest with the intention to replace fossil inputs, for example in European Commission communication since 2010.
However, @vivien2019HijackingBioeconomy point to the long history of biorefinery as a core aspect of this vision, which was discussed as early as the 1930s.
Hence, while the organization of the discourse into the labels biotechnology, bioecology and bioresource vision is a recent development, the underlying concepts have a richer history.
Over time these visions have not only coexisted, but also influenced each other by borrowing from, contributing to, and distancing themselves from the other visions of a bioeconomy [@allain2022BioeconomicsVsBioeconomy].
And although the exact role played by forests in a bioeconomy transition depends on the desired vision [@kleinschmit2014ShadesGreenSocial], there is no doubt that the Swedish vision places high demands on forests as primary providers of the required biomass. Or, as @belandlindahl2015SwedishForestryModel call it a forest sector that operates under a "more of everything" paradigm. Whether in the strategy formulated in @formas2012SwedishResearchInnovation, in Sweden's national forest program [@fischer2020SwedishForestsBioeconomy], or in public discourse, a small, well-connected group of actors center the bio-resource vision as legitimate and desirable [@holmgren2022WhoseTransformationThis].

Common ground in ambitions to realize a bioeconomy is the need for diverse types of social, technological and organizational innovation [@el-chichakli2016PolicyFiveCornerstones].  According to @jankovsky2021InnovationsForestBioeconomy there are four types of innovation crucial for a forest-based bioeconomy: (i) substitute products, (ii) new (bio-based) processes, (iii) new (bio-based) products, (iv) new behavior.
The need for innovation also features large in Swedish bioeconomy discourse and narratives therein [@bauer2017TechnologicalInnovationSystems; @bennich2021BiobasedEconomy2030], as, "\[t\]echnological innovation is key to a greener future" [@holmgren2022WhoseTransformationThis 42].
In a survey study sent to private organizations that received EU funding, @lovric2020FactorsDevelopmentInnovations find that the main type of European forest-based bioeconomy innovations are aimed at production processes, followed by goods and services, then business models.

In summary, despite -- or perhaps because of -- bioeconomy lacking a strong definition, the term is used to refer to different concepts, with different aims and objectives, foci, normative assumptions and hence implications for society and environment [@bugge2016WhatBioeconomyReview]. The summarized visions express themselves as narratives, actor coalitions and discourse artefacts in attempts to provide normative directionality to the transition.
This normative directionality is expected to translate into positive directionality of technological change through innovation.

@@

Operationalizing directionality suffers from frequent lack of differentiation of positive and normative directionality, in addition to inherent issues in the measurement of innovation.
Notable exceptions are the work on socio-technical system morphology by @andersson2021OutcomesDirectionalityMorphology, who identify changes in the configuration of three different renewable energy systems, and by @yang2022ShapingDirectionalitySustainability who use interviews and secondary source analysis to understand changes to insitutions in solar photovoltaic production in two Chinese provinces.
However, both studies deal more with the aggregate effects of directionality on socio-technical systems rather than with positive directionality of innovation directly.

Normative directionality can and has been assessed through analysis of discourse [@] and narratives [@bauer2018NarrativesBiorefineryInnovation], visions and aims expressed in strategy documents [@parks2022DirectionalityTransformativeInnovation], and R&D expenditure [@cappellano2022WhereInnovationMeets; @wydra2020MeasuringInnovationBioeconomy].
Where positive directionality of technological change has been assessed, scientific publications [@wydra2020MeasuringInnovationBioeconomy; @jankovsky2021InnovationsForestBioeconomy], and more often, patents are used [e.g., @aghion2016CarbonTaxesPath;@popp2005LessonsPatentsUsing;@moscona2022DoesDirectedInnovation].
Such data, however, suffer from limitations when attempting to assess the directionality of socio-technical systems.
While measures of normative directionality are equipped to identify the desired direction by different actors, visions need to first be accepted by innovators and reformulated into search.
R&D funding can help in this task, but data on R&D funding is often too broad, targeting entire sectors or specific industries [@popp2005LessonsPatentsUsing].
While patents are more granular and a closer proxy of technological change, they suffer from well documented measurement errors [@popp2005LessonsPatentsUsing]; not all innovations are patented, and not all patents become innovations [@arundel1998WhatPercentageInnovations].

Literature-based innovation output measures remedy these issues, by capturing commercialized innovations [@vanderpanne2007IssuesMeasuringInnovation].
Applying this method to  editor-reviewed articles published in 15 independent trade journals covering technological developments for general and highly specialized audiences, the SWINNO database contains `{python} n_swinno` innovations commercialized by Swedish companies between `{python} year_min` and `{python} year_max` [@sjoo2014DatabaseSwedishInnovations].
Since this method relies on the judgment of editiors and journalists, it can be described as significant innovations, omitting small reconfigurations of exisiting products and processes.
The SWINNO database has significant benefits compared to other innovation measures.
For one, it ensures that actual innovation, meaning commercialized products and processes, and thus that its positive directionality is captured.
Second, because the source articles are preserved, it provides a rich source of data, frequently providing insights into the aims, motivations, advantages and challenges of the covered innovation, in addition to descriptions of what the innovation consists of, which allows to assess and quantify the direction of technological change.

To capture the forest-based bioeconomy the database was searched for innovation associated with core forest sectors and keywords describing forest products which may be used in sectors along the bioeconomy value chain based on @wolfslehner2016ForestBioeconomyNew.
The list of keywords and sectors is presented in the appendix (@sec-appendix-data), additionally reproducible code to generate the figures and tables in this paper can be found at [`github.com/pjkreutzer/swinno-bioeconomy-directionality`](https://github.com/pjkreutzer/swinno-bioeconomy-directionality).

@@

@@

```{python}
# | label: Number of Results
# | output: asis
